\documentclass[acmsmall,anonymous=false,nonacm]{acmart}

%% ---------------------------------------------------------------------------
%%  Verified Optimistic Extraction from the Calculus of Constructions
%%  to a Polymorphic Blame Calculus (System Fomega + blame)
%% ---------------------------------------------------------------------------

\usepackage{mathpartir}
\usepackage{listings}
\usepackage{booktabs}
\usepackage{xcolor}

\settopmatter{printacmref=false}
\setcopyright{none}

%% --- notation macros -------------------------------------------------------
\newcommand{\dyn}{\ensuremath{\mathsf{Dyn}}}
\newcommand{\kstar}{\ensuremath{\star}}
\newcommand{\karr}{\ensuremath{\Rightarrow}}
\newcommand{\Prop}{\ensuremath{\mathsf{Prop}}}
\newcommand{\Set}{\ensuremath{\mathsf{Set}}}
\newcommand{\Kind}{\ensuremath{\mathsf{Kind}}}
\newcommand{\Nat}{\ensuremath{\mathsf{Nat}}}
\newcommand{\srt}{s}
\newcommand{\cast}[4]{\langle #2 \Rightarrow #3 \rangle^{#4}\, #1}
\newcommand{\gnd}[2]{\mathsf{gnd}(#1, #2)}
\newcommand{\isgnd}[2]{\mathsf{is\_gnd}(#1, #2)}
\newcommand{\blame}[1]{\mathsf{blame}\,#1}
\newcommand{\nub}[4]{\nu #1{:}#2 := #3.\,#4}
\newcommand{\ext}{\mathcal{E}}
\newcommand{\extt}{\mathcal{T}}
\newcommand{\extk}{\mathcal{K}}
\newcommand{\extc}{\mathcal{C}}
\newcommand{\coerce}[3]{\mathsf{coerce}\,#1\,#2\,#3}
\newcommand{\islarge}{\mathsf{large}}
\newcommand{\nf}{\mathsf{nf}}
\newcommand{\dyntoken}{\mathsf{dyn}_0}
\newcommand{\lint}{\ell_{\mathsf{int}}}
\newcommand{\reduces}{\longrightarrow}
\newcommand{\reducesstar}{\longrightarrow^{*}}
\newcommand{\conv}{\simeq}
\newcommand{\sepsim}{\mathrel{\lesssim}}
\newcommand{\hastype}[3]{#1 \vdash #2 : #3}
\newcommand{\wf}[1]{#1\ \mathsf{wf}}

%% --- listing style for CoC surface syntax ----------------------------------
\definecolor{kw}{rgb}{0.15,0.15,0.6}
\definecolor{cmt}{rgb}{0.3,0.5,0.3}
\lstdefinelanguage{coc}{
  morekeywords={fun,forall,Set,Prop,Kind,Axiom,Definition,Infer,Check,Extract,let,in},
  sensitive=true,
  morecomment=[s]{(*}{*)},
  moredelim=[s][\color{cmt}]{(*}{*)}
}
\lstset{
  language=coc,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{kw}\bfseries,
  commentstyle=\color{cmt}\itshape,
  columns=fullflexible,
  keepspaces=true,
  showstringspaces=false,
  breaklines=true,
  xleftmargin=1em
}

\begin{document}

\title{Extracting Blame}
\subtitle{Verified Optimistic Extraction from CoC to System~F$\omega$ + Blame}

\author{Joomy Korkut}
\email{jkorkut@bloomberg.net}
\orcid{0000-0001-6784-7108}
\affiliation{%
  \institution{Bloomberg}
  \city{New York}
  \state{NY}
  \country{USA}
}

\renewcommand{\shortauthors}{Korkut}

\begin{abstract}
Program extraction turns a dependently typed development into executable code,
but standard extraction erases dependent structure into ML-like programs whose
types cannot express the original invariants---and, where the erased structure
exceeds ML polymorphism, the extractor bridges the gap with unchecked casts
(\texttt{Obj.magic}), so a well-typed call from client code can crash the
extracted program with no catchable error. We ask a complementary question:
how much of a dependently typed program's structure can be preserved if we
extract into a \emph{typed} target, and can the translation be proved correct?

We present a mechanized extraction from the Calculus of Constructions (CoC) to
System~F$\omega$ extended with a dynamic type, explicit casts, and blame
tracking, together with the metatheory of both the target calculus and the
translation. The translation is \emph{optimistic}: it preserves the
parametric-polymorphic skeleton of CoC terms---including higher-kinded type
families---erases term indices from target types, and routes the genuine
residue of dependency (values used as types) through the dynamic type under a
reserved internal blame label. In Rocq, with no proof holes or project-local
axioms, we prove: for the target, kind-regular typing, operational
determinism, type-level confluence, progress, and preservation; for the
translation, well-typed extraction, simulation of source reduction by a
syntactic target relation, derivation independence, an exact syntactic
characterization of when extracted types avoid the dynamic type, and a
guarantee that an extracted term reduced in isolation never raises blame at an
external label. The simulation relation is syntactic rather than contextual,
and proof erasure is out of scope; we delimit both boundaries precisely. An
extracted OCaml REPL with verified de~Bruijn-to-named printing makes the
translation inspectable on eighteen dependently typed programming idioms.
\end{abstract}

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003790.10011740</concept_id>
<concept_desc>Theory of computation~Type theory</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
<concept_desc>Software and its engineering~Functional languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}
\ccsdesc[500]{Theory of computation~Type theory}
\ccsdesc[300]{Software and its engineering~Functional languages}

\keywords{program extraction, gradual typing, blame, dependent types,
Calculus of Constructions, System F$\omega$, mechanized metatheory, Rocq}

\maketitle

%% ==========================================================================
\section{Introduction}
%% ==========================================================================

Dependently typed languages such as Rocq (formerly Coq), Agda, Lean, and Idris
let a programmer state and prove precise properties of code. To \emph{run} that
code, one erases the static, non-computational parts and emits an executable
residue. Rocq's extraction \cite{letouzey2003extraction,letouzey2004these} is
the canonical example: it maps Gallina to OCaml, Haskell, or Scheme, erasing the
proof/type structure that the target languages cannot represent directly. This is enormously useful, but it makes two concessions.
First, the \emph{target type system cannot express the source's dependent
invariants}: a well-typed dependent program becomes an ML term whose safety is
no longer captured by the target's types. Second, the extractor is part of the
\emph{trusted computing base}: its correctness is argued informally, not proved
inside the proof assistant. Recent work by Forster, Sozeau, and
Tabareau~\cite{forster2024verified} verifies extraction from Coq to OCaml and
so addresses the second concession; our work is complementary and addresses
the first, studying extraction into a typed target that makes the lost
dependency explicit.

The first concession is sharper than ``the target types say less.''  Wherever
the dependent structure exceeds what ML polymorphism can express, Rocq's
extractor bridges the gap with \texttt{Obj.magic}, OCaml's unchecked cast.
This is not a corner case: we ported the eighteen showcased example programs
of \S\ref{sec:examples} to idiomatic CIC and ran Rocq~9.0's stock extractor,
and seven of the eighteen---bounds-checked indexing, heterogeneous lists,
type-safe \texttt{printf}, higher-kinded typeclass records, an intrinsically
typed interpreter, its computed-return-type variant, and a Tarski
universe---extract with \texttt{Obj.magic} casts or \texttt{Obj.t}-typed
interfaces (Table~\ref{tab:examples}).  The consequence is a runtime-safety
problem, not a cosmetic one.  The erased indices are exactly what made the
source code safe, so the extracted module has \emph{well-typed} OCaml calls
that violate the erased invariants; because the extractor hid the invariant
behind an unchecked cast, such a call is undefined behavior.  Calling the
extracted bounds-checked lookup out of bounds---a call OCaml's type checker
accepts---raises no exception; it segfaults (\S\ref{sec:examples}).  The
failure cannot be caught at runtime, and nothing attributes it to the
boundary that was violated.  A blame calculus replaces exactly this failure
mode: the boundary becomes a \emph{typed} cast whose failure is
$\blame{\ell}$---an ordinary, observable outcome licensed by target progress
and preservation (\S\ref{sec:target}) and attributed by its label to the
boundary that was crossed.  Runtime safety at the boundary becomes a theorem
of the target language rather than a discipline demanded of every client.

The idea of extracting a non-dependent polymorphic program from a CoC
development goes back to Paulin-Mohring's extraction of F$\omega$ programs from
proofs in the Calculus of Constructions~\cite{paulinmohring1989fomega}.  That
work uses a realizability interpretation and a distinction between informative
and non-informative propositions to remove logical content.  We return to the
same high-level source/target tension, but with a different target and theorem:
rather than extracting only the F$\omega$-realizable fragment, we extract
optimistically into F$\omega$ extended with $\dyn$, casts, blame, and
$\nu$-style sealing, so that the translation's fallback cases are explicit and
covered by an external-label non-generation theorem for isolated extraction.

This paper investigates this design point. We fix an expressive
\emph{typed} target and ask how much of a dependently typed program's structure
we can \emph{preserve}, and whether the translation can be \emph{verified}. Our
source is the Calculus of Constructions (CoC) \cite{coquand1988coc}, presented
as a pure type system (\S\ref{sec:source}) and mechanized following Barras'
\emph{Coq-in-Coq}
development \cite{barras1999,barras1997coqincoq}. Our target is System~F$\omega$
extended with a dynamic type $\dyn$, type-directed casts, and blame tracking:
a \emph{polymorphic blame calculus} in the line of Wadler and Findler
\cite{wadler2009blame} and Ahmed, Findler, Siek, and Wadler
\cite{ahmed2011blame}. We call it \emph{System~F$\omega$ + blame}.

\paragraph{Why an explicit dynamic/blame target?}
F$\omega$ preserves a large parametric-polymorphic skeleton: type
abstraction, type application, and higher-kinded type families
\cite{reynolds1983types}. CoC has strictly more---full dependency, in which
types may mention terms. The mismatch is fundamental for a direct,
type-preserving translation into ordinary F$\omega$: arbitrary term dependency is
not part of the target's type formation. Adding $\dyn$ and explicit casts
represents this mismatch without
falling back to a fully untyped image: we keep everything F$\omega$ can express
and route the genuine residue of dependency through the dynamic type $\dyn$,
with casts and a reserved failure label marking the boundary. We call the
translation \emph{optimistic}: it never protectively dynamizes structure the
target can express, and a recursively defined source predicate characterizes
exactly when the extracted type contains no $\dyn$
(Theorem~\ref{thm:dynfree}). We do not prove that this predicate is
semantically maximal.

Concretely, three things happen at the boundary (\S\ref{sec:overview}
illustrates all three):
\begin{itemize}
\item \textbf{Term-indexed dependencies are erased from target types.}
A length-indexed vector \texttt{Vec A n} becomes the plain polymorphic type
\texttt{Vec A}: the target type no longer records the relation between the
vector and the term index~$n$.  The term~$n$ itself is not necessarily erased:
if the source program passes it to a function or eliminator, it remains an
ordinary target term argument.  What disappears is the dependent typing
invariant, not every runtime occurrence of the index.
\item \textbf{Type indices are kept.} A type family applied to a type argument
extracts to a type application, preserving parametricity and the family's
higher-kinded structure.
\item \textbf{Values used as types become dynamic residue.}
A source term used where the target expects a type, such as the decoding
function of a Tarski universe, has no direct F$\omega$ type-level image.
In target types this residue becomes $\dyn$; in term position, extraction
uses a reserved internal blame term cast through $\dyn$. It is well typed at
every proper type and the syntactic relation connects it to any term, but it is
a run-time error by design: it marks a fallback selected by this translation.
\end{itemize}

\paragraph{Artifact scope.}
The contribution is not the observation that CoC contains an F$\omega$-extractable fragment:
Paulin-Mohring established that lineage~\cite{paulinmohring1989fomega}.  Nor is
the novelty verified erasure to executable code: MetaCoq/MetaRocq, CertiCoq,
\OE{}uf, and Forster et al.'s verified extraction address that
space~\cite{sozeau2020metacoq,anand2017certicoq,mullen2018oeuf,forster2024verified}.
The artifact combines three choices: (1)~an explicit polymorphic
blame target with F$\omega$-style type computation, $\dyn$, casts, and
$\nu$-sealing; (2)~an optimistic extraction that preserves the
F$\omega$-expressible skeleton and marks the rest with internal blame; and
(3)~a Rocq mechanization proving target progress and preservation, well-typed
extraction, source-reduction simulation, derivation independence,
external-label non-generation, operational determinism, and type-level
confluence.

\paragraph{Contributions.}
\begin{itemize}
\item We mechanize an F$\omega$-style \emph{polymorphic blame calculus}
(\S\ref{sec:target}): syntax, a small-step semantics proved \emph{deterministic},
a kinding relation, type-level $\beta$/definitional equality with
\emph{confluence} and \emph{Church--Rosser}, a typing conversion rule, a typing
judgment, subtyping, a syntactic simulation relation, and label-safety
invariants.  We further
develop the target's structural metatheory---term and type weakening, term and
type substitution, inversion through conversion, and canonical forms---all in
Rocq with no admitted lemmas or project-local axioms; the assumption set is
audited separately (\S\ref{sec:mech}).  This is a
target variant inspired by Blame for All~\cite{ahmed2011blame}, with $\dyn$
restricted to kind $\kstar$. Its typing constructors carry the kinding and
ground-tag premises needed for regularity. We prove annotation regularity,
progress, one-step preservation, and multi-step preservation.

\item We define an \emph{optimistic, conversion-invariant} extraction from CoC
to this target (\S\ref{sec:extraction}).  The design returns to
Paulin-Mohring's CoC-to-F$\omega$ insight~\cite{paulinmohring1989fomega} but
adds an explicit dynamic/blame boundary: the F$\omega$-expressible skeleton is
preserved, while cases not represented statically by this translation use
$\dyn$ and internal blame.  The classification of binders as type- or term-level
uses a semantic, reduction-stable predicate ($\islarge$)---not a syntactic
approximation---and types are normalized before translation, so the translation
respects definitional equality. The extraction recurses directly on \emph{typing
derivations}, which required moving CoC's entire reduction, conversion, and
typing infrastructure from \texttt{Prop} to \texttt{Type}.

\item We prove the metatheory (\S\ref{sec:metatheory}): well-typedness
preservation of the extraction; one-step and multi-step \emph{simulation} of
CoC's canonical reduction relations by the target's syntactic relation;
independence of the extraction from the choice of typing derivation; a
syntactic lemma relating a concrete instantiation to the $\dyn$ instantiation;
external-label non-generation for isolated extracted terms; and an
\emph{optimism} result characterizing a source-type fragment whose extraction
contains no $\dyn$.

\item We build a REPL (\S\ref{sec:repl}) that type-checks CoC terms and prints
their F$\omega$ + blame extractions, with a \emph{verified} de~Bruijn-to-named
renderer whose name recovery is capture-free by construction and whose checked
OCaml wrapper validates namespace and scope preconditions.

\item We present eighteen showcased command batches (\S\ref{sec:overview},
\S\ref{sec:examples}) that make visible what the extraction keeps, erases, and
dynamizes. Axioms provide only the inductive kit---type formers, constructors,
and elimination/induction principles---and the deliberate value-as-type
boundaries; all other functions are ordinary definitions built from the
eliminators.
\end{itemize}

The entire development contains no admitted lemmas or project-local axioms;
the audited core theorem set uses only the standard-library
\texttt{eq\_rect\_eq} (\S\ref{sec:mech}).

\paragraph{Scope.}
We study a single design point: verified, typed, optimistic extraction from a
pure-PTS source to a polymorphic blame target. Two boundaries deserve emphasis
up front (\S\ref{sec:limitations} catalogues the rest). First, the extraction
is defined as a total mathematical function on derivations, but it invokes
normalization and a type-checking decision procedure, so it is a metatheoretic
object rather than an efficient compiler pass. Second, our simulation is a
syntactic bookkeeping relation, not contextual approximation. We also do not
address proof erasure, inductive families, universe polymorphism, or verified
production compilation; each marks a possible extension, not a limitation of
the results we do prove. Our aim is to chart the design space of \emph{typed,
verified} extraction, not to ship a production back end.

%% ==========================================================================
\section{Overview by Example}\label{sec:overview}
%% ==========================================================================

We begin with the phenomenology. Throughout, axioms introduce only what native
inductive definitions would provide---type formers, constructors, and
elimination principles---the honest way to model inductive families in a bare
pure type system. Every other function is an ordinary \texttt{Definition}
built from the eliminators (the REPL expands definitions at use sites,
\S\ref{sec:repl}), and its extraction is governed by our verified
translation. Each output comment below is taken
from the corresponding REPL golden file (\S\ref{sec:repl}), with line breaks
added for readability. The
printer renders extracted terms in Rocq-like concrete syntax: term and type
abstractions both print as \texttt{fun} with consecutive binders grouped,
quantifiers print as \texttt{forall}, type applications as $[\cdot]$, casts as
\texttt{cast<A => B>\^{}$\ell$(e)}, and the dynamic type as $\dyn$.

\paragraph{Term-indexed dependencies disappear from target types.}
A bounds-checked lookup indexes a length-indexed vector by a provably in-range
position:
\begin{lstlisting}
Axiom Vec : Set -> Nat -> Set.
Axiom Fin : Nat -> Set.

Definition vnth (A : Set) (n : Nat) (xs : Vec A n) (i : Fin n) : A :=
  vrec A (fun (m : Nat) => Fin m -> A)
    (fun (i0 : Fin NZ) => fin0 A i0)
    (fun (m : Nat) (a : A) (rest : Fin m -> A) (j : Fin (NS m)) =>
     fin_case m A a rest j)
    n xs i.

Extract vnth.
(* ==> fun (A : *) (n : Nat) (xs : Vec A) (i : Fin) =>
         vrec [A] [Fin -> A]
           (fun (x0 : Fin) => fin0 [A] x0)
           (fun (x0 : Nat) (x1 : A) (x2 : Fin -> A) (x3 : Fin) =>
              fin_case x0 [A] x1 x2 x3)
           n xs i *)
\end{lstlisting}
\texttt{vnth} is defined by vector recursion at the motive
$P\,m = \texttt{Fin}\,m \to A$, refuting the index at the empty vector
(\texttt{fin0}) and case-splitting it at a cons (\texttt{fin\_case}).
The length~$n$ is a term; it appears in the source types \texttt{Vec A n} and
\texttt{Fin n} but is dropped from the extracted types \texttt{Vec A} and
\texttt{Fin}, while the implementation survives as an ordinary recursion.
The term arguments \texttt{n} and \texttt{i} themselves remain
because \texttt{vnth} uses them as ordinary data.  What the target no longer
records is the dependent invariant that \texttt{i} is bounded by~$n$: the
typing relationship is what disappears, not every runtime occurrence of the
index.

\paragraph{Equality indices vanish from types; proof terms remain.}
Propositional equality lives in \texttt{Prop}. The current extractor does not
implement Coq-style proof erasure, but it erases the equality endpoints from
target types: \texttt{Eq A x y} becomes \texttt{Eq A}. Proof arguments and
eliminators such as \texttt{eq\_rect} remain as ordinary residual terms:
\begin{lstlisting}
Extract fun (A : Set) (x y : A) (h : Eq A x y) =>
  eq_rect A x (fun (z : A) => Eq A z x) (refl A x) y h.
(* ==> fun (A : *) (x y : A) (h : Eq A) =>
         eq_rect [A] x [Eq A] (refl [A] x) y h *)
\end{lstlisting}
In a dependently typed language, \emph{transport} (\texttt{eq\_rect}) is how one
crosses between two provably equal types; it is the static ancestor of a cast.
This example demonstrates index erasure from types, not proof erasure.
A future optimization could erase proof terms more aggressively.

\paragraph{Dependent arities collapse.}
Type-safe \texttt{printf} indexes a format by the type of the variadic function
it denotes; the number and types of arguments are computed from the format
value. The dependency is entirely static:
\begin{lstlisting}
Check sprintf (Str -> Nat -> Str) (fstr (Nat -> Str) (fint Str fstop)).
(* ==>  Inferred type: Str -> Nat -> Str *)
\end{lstlisting}
Here the computed arity \texttt{Str -> Nat -> Str} is a perfectly ordinary
F$\omega$ arrow type; the dependent arity leaves no dynamic checks.
\texttt{sprintf} itself is a definition---an accumulator-style fold over the
format at motive $P\,X = \texttt{Str} \to X$---and its extraction inlines
that fold at an ordinary polymorphic target type.

\paragraph{Where the dynamic type earns its keep.}
The examples above stay within the target's static skeleton: their dependent
indices are erased from target types without introducing~$\dyn$. The exception
is a value used \emph{as a
type}. A Tarski universe has codes \texttt{U} and a decoder \texttt{El : U ->
Set}; the function \texttt{fun c => El c} maps data to types:
\begin{lstlisting}
Extract fun (c : U) => El c.
(* ==>  fun (c : U) => cast<Dyn => U -> Dyn>^2+(blame(2+)) c *)
\end{lstlisting}
This is the boundary. \texttt{El} has no F$\omega$ image, so the extraction
emits a coercion-failure blame, cast through the dynamic type~$\dyn$, and
applies it to the code. The superscript $2{+}$ is the blame label: identifier
$2$ is the reserved internal ``no F$\omega$ image'' marker (defined
in~\S\ref{sec:target}), and $+$ is its polarity. This is the translation's
explicit fallback for ``a value in type position,'' and our external-label
theorem
(\S\ref{sec:metatheory}) proves that this isolated extracted term cannot reduce
to an external label. It is not a theorem about arbitrary linking contexts.

\medskip
The lesson of these examples, and of the eighteen showcased batches in
\S\ref{sec:examples}, is that an explicit dynamic/blame target lets a single
total mathematical translation preserve the polymorphic skeleton, remove many dependent indices from target
types, and mark the residue of dependency syntactically with $\dyn$ and a
reserved failure label.

%% ==========================================================================
\section{The Source: the Calculus of Constructions}\label{sec:source}
%% ==========================================================================

Our source is the Calculus of Constructions, presented as a pure type system
(PTS) \cite{barendregt1991lambda} with three sorts. We reuse Barras'
\emph{Coq-in-Coq} formalization \cite{barras1999} for terms, de~Bruijn
substitution, reduction, confluence, strong normalization, and decidable
type-checking, adapting it as described in \S\ref{sec:mech}.

\paragraph{Syntax and sorts.}
Terms are the usual PTS terms over de~Bruijn indices:
\[
t, T ::= s \mid n \mid \lambda T.\,t \mid t\ t \mid \Pi T.\,T,
\qquad
s ::= \Prop \mid \Set \mid \Kind.
\]
The sort axioms are $\Prop : \Kind$ and $\Set : \Kind$; the top sort $\Kind$ is
untypable, which is what averts Girard's paradox \cite{girard1972}. The product
rule is unrestricted over the typable sorts: for all $s_1, s_2$,
\[
\inferrule{\hastype{\Gamma}{T}{s_1} \\ \hastype{\Gamma, T}{U}{s_2}}
          {\hastype{\Gamma}{\Pi T.\,U}{s_2}}
\]
so products are impredicative in both $\Prop$ and $\Set$. Figure~\ref{fig:coc}
gives the typing rules in full. Conversion is definitional: the rule
\textsc{Conv} lets a term inhabit any type convertible ($\conv$, the
reflexive-symmetric-transitive closure of $\beta$-reduction) to its inferred one.

\begin{figure}
\small
\begin{mathpar}
\inferrule*[right=\textsc{Prop}]{\wf{\Gamma}}{\hastype{\Gamma}{\Prop}{\Kind}}
\and
\inferrule*[right=\textsc{Set}]{\wf{\Gamma}}{\hastype{\Gamma}{\Set}{\Kind}}
\and
\inferrule*[right=\textsc{Var}]{\wf{\Gamma} \\ \Gamma(n) = T}{\hastype{\Gamma}{n}{T}}
\\
\inferrule*[right=\textsc{Lam}]
  {\hastype{\Gamma}{T}{s_1} \\ \hastype{\Gamma, T}{U}{s_2} \\ \hastype{\Gamma, T}{M}{U}}
  {\hastype{\Gamma}{\lambda T.\,M}{\Pi T.\,U}}
\and
\inferrule*[right=\textsc{App}]
  {\hastype{\Gamma}{u}{\Pi V.\,U_r} \\ \hastype{\Gamma}{v}{V}}
  {\hastype{\Gamma}{u\ v}{U_r[v]}}
\\
\inferrule*[right=\textsc{Prod}]
  {\hastype{\Gamma}{T}{s_1} \\ \hastype{\Gamma, T}{U}{s_2}}
  {\hastype{\Gamma}{\Pi T.\,U}{s_2}}
\and
\inferrule*[right=\textsc{Conv}]
  {\hastype{\Gamma}{t}{U} \\ U \conv V \\ \hastype{\Gamma}{V}{s}}
  {\hastype{\Gamma}{t}{V}}
\end{mathpar}
\caption{Typing for the source Calculus of Constructions (de~Bruijn; $\Gamma(n)$
denotes the appropriately lifted $n$-th binding). Well-formedness $\wf{\Gamma}$
threads through the rules.}
\label{fig:coc}
\Description{Typing rules for the source Calculus of Constructions.}
\end{figure}

\paragraph{The large/small distinction.}
Extraction hinges on separating \emph{type-level} from \emph{term-level}
objects. A source object is \emph{large} when its type is the top sort $\Kind$:
\[
\islarge(\Gamma, T) \;\triangleq\; \hastype{\Gamma}{T}{\Kind}.
\]
Large things are kinds---$\Set$, $\Prop$, $\Nat \to \Set$, and so on---which
classify types; they must become F$\omega$ type-level binders and quantifiers.
Small things (sort $\Prop$ or $\Set$) classify terms and become term-level
binders. The crucial property, which we prove (\S\ref{sec:metatheory}), is that
$\islarge$ is \emph{stable under reduction}: subject reduction guarantees that
if $T$ has sort $\Kind$ and $T \reduces T'$, then $T'$ has sort $\Kind$. This is
what makes every classification decision in the extraction respect definitional
equality. A purely syntactic approximation ($\mathsf{classifier}$, used only in
positions already known to be normal) is sound but not reduction-stable, and we
are careful never to base a conversion-sensitive decision on it.

%% ==========================================================================
\section{The Target: an F$\omega$-Style Blame Calculus}\label{sec:target}
%% ==========================================================================

The target follows the polymorphic blame calculus of Ahmed et
al.~\cite{ahmed2011blame}, lifted from System~F to F$\omega$-style
higher-kinded type syntax and extended with a dynamic type $\dyn$,
type-directed casts, and blame. We give a machine-checked
formalization of this F$\omega$-style blame variant; the full development lives in the
\texttt{BlameFOmega} directory. The type language has the essential F$\omega$ structure: it has
type operators ($\Lambda$/juxtaposition), a kinding relation, a type-level
$\beta$ rule with its definitional-equality closure, and a typing conversion
rule. The term judgment embeds the kinding and ground-tag premises required by
its annotations. We prove result-type regularity, annotation regularity,
progress, and preservation for the full reduction relation.

\paragraph{Kinds, types, terms.}
\begin{align*}
K &::= \kstar \mid K \karr K & \text{(kinds)}\\
A, B &::= \alpha \mid A \to B \mid \forall \alpha{:}K.\,A \mid \Lambda \alpha{:}K.\,A \mid A\ B \mid \dyn & \text{(types)}\\
e &::= x \mid \lambda x{:}A.\,e \mid e\ e \mid \Lambda \alpha{:}K.\,e \mid e\,[A] \\
  &\quad \mid \cast{e}{A}{B}{p} \mid \gnd{e}{G} \mid \isgnd{e}{G} \mid \blame p \mid \nub{\alpha}{K}{A}{e} & \text{(terms)}
\end{align*}
Types include the polymorphic quantifier $\forall$, the type-operator
abstraction/application $\Lambda$/juxtaposition (needed to preserve CoC's
higher-kinded type families), and the dynamic type $\dyn$. The dynamic type is a
proper type of kind $\kstar$; we do not include a kind-indexed family
$\dyn_K$. Terms extend F$\omega$ with: a cast $\cast{e}{A}{B}{p}$ that coerces
$e$ from $A$ to $B$ under blame label $p$; ground injection $\gnd{e}{G}$ and the
ground test $\isgnd{e}{G}$, which mediate between $\dyn$ and \emph{ground types}
$G ::= \dyn \to \dyn \mid N$, where $N ::= \alpha \mid N\,A$ ranges over
\emph{neutral} types (a type variable or a type variable applied to further
types)---the F$\omega$ generalization of Blame-for-All's type-variable ground
tag, needed once type applications, not just variables, can appear at the
$\dyn$ boundary; a blame term $\blame p$; and a
\emph{type-sealing} binder $\nub{\alpha}{K}{A}{e}$ that introduces a fresh
abstract type variable $\alpha$ of kind $K$ whose hidden implementation is $A$,
used to model polymorphic instantiation dynamically (it is the reduct of a type
application $\Lambda$-elimination). Unlike Blame for All's global store of
type names, our $\nu$ is a \emph{scoped} binder, eliminated by commuting rules
rather than extruded; a ground tag that mentions the sealed variable cannot
outlive its binder and instead blames the reserved label~$0$
(Figure~\ref{fig:target-redexes}). Sealed values therefore cannot escape
their $\nu$-scope inside $\dyn$; \S\ref{sec:limitations} discusses this
trade-off. \paragraph{Blame labels: which cast failed, and whose fault it is.}
A blame label answers the question: \emph{if a cast fails, who is to blame?}
A cast $\cast{e}{A}{B}{p}$ that cannot reconcile $A$ with $B$ does not get
stuck and does not silently return a wrong value---it reduces to $\blame p$.
The label $p$ is the cast's name tag, so a failure points at exactly
\emph{which} boundary broke. A label is a pair. Its \emph{identifier} names
the boundary; its \emph{polarity}---a single bit, written as a sign---records
\emph{which side} of that boundary is at fault: positive~($+$) blames the
value flowing through the cast, negative~($-$) blames the context that
consumed it. Following Wadler and Findler, a function cast pushes a cast onto
its argument with the polarity \emph{flipped} (the $\bar p$ in
Figure~\ref{fig:target-redexes}), because responsibility for the argument
belongs to the other party. Thus $\lint$ with positive polarity prints as
$\lint{+}$, or simply $2{+}$ using its identifier.

Identifiers split into two classes. Three small identifiers are
\emph{reserved} for the translation's own plumbing---failures only the
internal machinery can trigger, never a contract the source programmer wrote:
\begin{center}
\begin{tabular}{cl}
\toprule
id & raised when \ldots \\
\midrule
$0$ & a $\nu$-sealed type variable would escape its scope through $\dyn$ \quad($\nu$-tampering) \\
$1$ & the ground test $\isgnd{\cdot}{G}$ inspects a sealed value \quad($\mathsf{is\_gnd}$-tampering) \\
$2$ & the extraction meets a source construct with no F$\omega$ image \quad(coercion failure; written $\lint$) \\
\bottomrule
\end{tabular}
\end{center}
Every identifier $\ge 3$ is \emph{external}: it names a genuine boundary that
a client of the calculus could introduce. Keeping the three internal classes
distinct---rather than collapsing them into one ``internal error''---preserves
their diagnostic origin, and it is exactly what gives the safety theorem
(\S\ref{sec:metatheory}) its bite: an extracted term reduced in isolation can
only ever raise an \emph{internal} label ($0$, $1$, or $2$), a mark the
translation itself planted, and \emph{never} an external one.

\paragraph{Executable compatibility.}
Unlike Blame for All's System-F \emph{consistency} relation, our $\mathsf{compat}(A,B)$
is an \emph{executable} judgment: every constructor corresponds directly to a
cast-reduction rule, so a well-typed cast always has an operational step and
there is no catch-all ``stuck cast'' case. Except for identity casts, which
step without inspecting their annotation, each constructor decomposes the cast
on a canonical head. This is one of the places where the F$\omega$ extension is not routine: in
System~F, Blame-for-All compatibility can be read as both consistency and a cast
strategy, but once type-level functions and neutral type-family applications are
present, castability must be made executable---every compatibility constructor
becomes a cast-elaboration rule. $\mathsf{compat\_refl}$
licenses the identity cast; $\mathsf{compat\_arrow}$ (contravariant domain,
covariant codomain) licenses the structural WRAP step; $\mathsf{compat\_all}$
licenses the analogous structural step under a matching $\forall$-binder;
$\mathsf{compat\_generalize}$ and $\mathsf{compat\_instantiate}$ license casting
into and out of a $\forall$; $\mathsf{compat\_to\_dyn}$ licenses injection at a
ground tag, itself requiring $\mathsf{compat}(A,G)$ so injection only ever
manufactures a cast the rest of the judgment already accepts; and
$\mathsf{compat\_from\_dyn}$ licenses projecting a ground value back out of
$\dyn$. The judgment is decidable ($\mathsf{compat\_dec}$), which is what makes
$\coerce{\cdot}{\cdot}{\cdot}$ (\S\ref{sec:extraction}) a total function.
Compatibility itself is syntactic and context-free, whereas typing also admits
context-indexed definitional equality. In particular,
$\mathsf{compat\_refl}$ accepts identical non-normal annotations because the
identity step does not inspect them; we do not claim coherence of
$\mathsf{compat}$ modulo definitional equality. Non-identity clients normalize
annotations before selecting a compatibility derivation.

\paragraph{Semantics, typing, and safety.}
We define a small-step reduction $e \reduces e'$ (with values, $\beta$/type-$\beta$
reduction, cast reductions that split structural casts and fire blame on ground
mismatch, and $\nu$-sealing), a kinding relation
$\mathsf{wf\_typ}(\Delta, A, K)$, and a typing judgment
$\hastype{\Delta}{e}{A}$ over contexts $\Delta$ binding term variables (at a
type), type variables (at a kind), and type-definition bindings introduced by
$\nu$-sealing. The dynamic type is inhabited by ground
injections, and casts are well typed between compatible, well-kinded types. The
rules for abstractions, type application, casts, ground wrappers/tests, blame,
$\nu$, and conversion carry the premises needed for kind regularity. We
port the subtyping and \emph{safety} apparatus of \cite{ahmed2011blame}: a
predicate $\mathsf{safe}(p, e)$ ensures every cast in $e$ that could blame $p$
carries the right positive/negative subtyping evidence, and safety is preserved
by reduction. From this we obtain the central guarantee.

\begin{theorem}[External label-safety invariant; \texttt{blame\_theorem}, mechanized]
\label{thm:blame}
If $\mathrm{lbl\_id}(p) \ge 3$ and $\mathsf{safe}(p, e)$, then
$\neg\,(e \reducesstar \blame p)$.
\end{theorem}
In words: an extracted program may still fail, but only by pointing at its own
internal plumbing---labels $0$, $1$, or $2$---never at an external boundary
(any label $\ge 3$). This is the invariant theorem formalized by the artifact.
It is inspired by the safety development in \emph{Blame for All}, but it is not
presented as one of that paper's numbered contextual corollaries.

Figure~\ref{fig:target-typing} gives the target typing rules used by the
mechanization.  In the last rule, $\mathsf{defeq}_{\Delta}$ includes
F$\omega$ type conversion and equations exposed by $\nu$ definition bindings.

\begin{figure*}
\small
\begin{mathpar}
\inferrule*[right=\textsc{T-Var}]
  {\Delta(x)=A}{\hastype{\Delta}{x}{A}}

\inferrule*[right=\textsc{T-Abs}]
  {\mathsf{wf\_typ}(\Delta,A,\kstar) \\ \hastype{\Delta,x{:}A}{e}{B}}
  {\hastype{\Delta}{\lambda x{:}A.e}{A\to B}}

\inferrule*[right=\textsc{T-App}]
  {\hastype{\Delta}{e_1}{A\to B} \\ \hastype{\Delta}{e_2}{A}}
  {\hastype{\Delta}{e_1\ e_2}{B}}

\inferrule*[right=\textsc{T-TAbs}]
  {\hastype{\Delta,\alpha{:}K}{e}{A}}
  {\hastype{\Delta}{\Lambda\alpha{:}K.e}{\forall\alpha{:}K.A}}

\inferrule*[right=\textsc{T-TApp}]
  {\hastype{\Delta}{e}{\forall\alpha{:}K.A} \\
   \mathsf{wf\_typ}(\Delta,B,K)}
  {\hastype{\Delta}{e[B]}{A[B/\alpha]}}

\inferrule*[right=\textsc{T-Cast}]
  {\hastype{\Delta}{e}{A} \\ \mathsf{compat}(A,B) \\
   \mathsf{wf\_typ}(\Delta,A,\kstar) \\
   \mathsf{wf\_typ}(\Delta,B,\kstar)}
  {\hastype{\Delta}{\cast{e}{A}{B}{p}}{B}}

\inferrule*[right=\textsc{T-Gnd}]
  {\hastype{\Delta}{e}{G} \\ \mathsf{wf\_ground}(\Delta,G)}
  {\hastype{\Delta}{\gnd{e}{G}}{\dyn}}

\inferrule*[right=\textsc{T-IsGnd}]
  {\hastype{\Delta}{e}{\dyn} \\ \mathsf{wf\_ground}(\Delta,G)}
  {\hastype{\Delta}{\isgnd{e}{G}}{\dyn\to\dyn\to\dyn}}

\inferrule*[right=\textsc{T-Blame}]
  {\mathsf{wf\_typ}(\Delta,A,\kstar)}
  {\hastype{\Delta}{\blame p}{A}}

\inferrule*[right=\textsc{T-Nu}]
  {\hastype{\Delta,\alpha{:}K=A}{e}{B} \\
   \mathsf{wf\_typ}(\Delta,A,K)}
  {\hastype{\Delta}{\nub{\alpha}{K}{A}{e}}{B[A/\alpha]}}

\inferrule*[right=\textsc{T-Conv}]
  {\hastype{\Delta}{e}{A} \\
   \mathsf{defeq}_{\Delta}(A,B,\kstar) \\
   \mathsf{wf\_typ}(\Delta,B,\kstar)}
  {\hastype{\Delta}{e}{B}}
\end{mathpar}
\caption{Kind-regular target term typing (Rocq: \texttt{typing}).}
\label{fig:target-typing}
\Description{Typing rules for the target System F-omega blame calculus.}
\end{figure*}

Figure~\ref{fig:target-redexes} records the characteristic redexes. Evaluation
contexts are represented in Rocq by explicit congruence and blame-propagation
constructors; the figure omits those routine rules and abbreviates the side
conditions of executable compatibility.

\begin{figure*}
\small
\begin{mathpar}
(\lambda x{:}A.e)\ v \reduces e[v/x]

(\Lambda\alpha{:}K.v)[A] \reduces \nub{\alpha}{K}{A}{v}

\cast{v}{A}{A}{p} \reduces v

\cast{v}{A_1\to A_2}{B_1\to B_2}{p}
 \reduces \lambda x{:}B_1.
   \cast{v\,(\cast{x}{B_1}{A_1}{\bar p})}{A_2}{B_2}{p}

\cast{v}{\forall\alpha{:}K.A}{\forall\alpha{:}K.B}{p}
 \reduces \Lambda\alpha{:}K.\cast{v^\uparrow[\alpha]}{A}{B}{p}

\cast{v}{A}{\forall\alpha{:}K.B}{p}
 \reduces \Lambda\alpha{:}K.\cast{v^\uparrow}{A^\uparrow}{B}{p}

\cast{v}{\forall\alpha{:}\kstar.A}{B}{p}
 \reduces \cast{v[\dyn]}{A[\dyn/\alpha]}{B}{p}

\cast{v}{A}{\dyn}{p} \reduces \gnd{\cast{v}{A}{G}{p}}{G}
\quad(\mathsf{ground\_tag}(A,G))

\cast{\gnd{v}{G}}{\dyn}{A}{p} \reduces \cast{v}{G}{A}{p}
\quad(\mathsf{compat}(G,A))

\cast{\gnd{v}{G}}{\dyn}{A}{p} \reduces \blame p
\quad(\neg\mathsf{compat}(G,A))

\isgnd{\gnd{v}{G}}{G} \reduces \lambda x{:}\dyn.\lambda y{:}\dyn.\,x
\quad(\neg\mathsf{neutral}(G))

\isgnd{\gnd{v}{H}}{G} \reduces \lambda x{:}\dyn.\lambda y{:}\dyn.\,y
\quad(G \neq H,\ \neg\mathsf{neutral}(H))

\isgnd{\gnd{v}{H}}{G} \reduces \blame 1
\quad(\mathsf{neutral}(H))

\nub{\alpha}{K}{A}{\gnd{v}{G}} \reduces
 \gnd{\nub{\alpha}{K}{A}{v}}{G[A/\alpha]}
\quad(\alpha\notin G)

\nub{\alpha}{K}{A}{\gnd{v}{G}} \reduces \blame 0
\quad(\alpha\in G)
\end{mathpar}
\caption{Characteristic target redexes. Value, head-shape, ground-tag, and
compatibility side conditions are as in \texttt{semantics.v}; in particular
the injection rules require $A \neq \dyn$ and a non-$\forall$ head, so a
dynamic value reveals its ground tag before any $\forall$-generalization
fires. The $\mathsf{is\_gnd}$ test returns a Church boolean at $\dyn$; a
neutral tag signals sealing tampering and blames the reserved label~$1$.}
\label{fig:target-redexes}
\Description{The characteristic reduction rules of the target calculus.}
\end{figure*}

\paragraph{Determinism.}
The small-step relation is deterministic: we prove
$\mathsf{step\_deterministic}$, that $e \reduces e_1$ and $e \reduces e_2$
imply $e_1 = e_2$. The proof rests on facts we also mechanize: values,
variables, and $\blame p$ are irreducible ($\mathsf{value\_not\_step}$,
$\mathsf{var\_not\_step}$, $\mathsf{blame\_not\_step}$), and every type has at
most one ground tag ($\mathsf{ground\_tag\_functional}$), which makes the
ground-cast rules non-overlapping. Proving determinism flushed out one real
overlap---a cast from $\dyn$ to a universal type could fire both
$\mathsf{step\_generalize}$ and
$\mathsf{step\_collapse}$/$\mathsf{step\_conflict}$---which we fixed by
guarding $\mathsf{step\_generalize}$ with $A \neq \dyn$: a dynamic value
reveals its ground tag first.

\paragraph{Type-level equality and confluence.}
The target has genuine F$\omega$ type-level computation: a type-level
reduction $\mathsf{ty\_step}$ with the $\beta$ rule
$(\Lambda\alpha{:}K.\,A)\,B \reduces A[B]$ and congruences, its
definitional-equality closure $\mathsf{ty\_equiv}$, and a typing conversion
rule $\mathsf{typing\_conv}$ (with a $\mathsf{wf\_typ}$ premise). We mechanize
\emph{confluence} of $\mathsf{ty\_step}$ by the parallel-reduction (Takahashi)
method---the full de~Bruijn substitution-lemma chain for $\mathsf{typ}$, a
complete-development function, the triangle lemma, the diamond property, and
Church--Rosser for $\mathsf{ty\_equiv}$---and derive the head-constructor
inversions ($\mathsf{ty\_equiv}$ is injective on $\to$ and $\forall$ and keeps
$\dyn$/$\to$/$\forall$/variables distinct). We also make the
$\dyn$-instantiation rules kind-correct: since $\dyn$ lives only at kind
$\kstar$, the reduction/subtyping/compatibility rules that substitute $\dyn$
into a quantifier are restricted to $\forall\alpha{:}\kstar$.

\paragraph{Structural metatheory.}
On top of confluence we mechanize the structural typing lemmas: term and type
weakening (with the two-namespace context-shift bookkeeping and
$\mathsf{tlift}$-preservation of $\mathsf{compat}$ and $\mathsf{ty\_equiv}$),
term substitution ($\mathsf{typing\_subst}$) and neutral type substitution
($\mathsf{typing\_tsubst}$, which requires the substituted type to be
$\mathsf{neutral}$, including $\mathsf{wf\_typ\_tsubst}$ and
$\mathsf{compat\_tsubst}$), inversion through $\mathsf{typing\_conv}$
(recovering a derivation's shape up to conversion for each term former), and
canonical forms (a value at an arrow/$\forall$ type is a
$\lambda$/$\Lambda$).

\paragraph{Progress and preservation.}
Building on the structural metatheory and on executable $\mathsf{compat}$, we
prove $\mathsf{progress}$: every closed term well typed under the kind-regular
judgment is a value, steps, or is $\blame p$ for some $p$. The cast case is
exactly the case analysis on $\mathsf{compat}(A,B)$: every constructor of the
executable judgment corresponds to a concrete reduction rule (ID, WRAP, the
structural $\forall/\forall$ step $\mathsf{step\_all\_all}$, GENERALIZE,
INSTANTIATE, GROUND, COLLAPSE/CONFLICT), so there is no leftover ``stuck
cast'' shape to handle by fiat---progress falls directly out of
$\mathsf{compat\_dec}$. The $\mathsf{is\_gnd}$/$\nu$-tamper detection is
generalized from Blame-for-All's bare type-variable case to \emph{any} neutral
ground tag $N$: a sealed variable can hide inside $N\,A$, not just a bare
$\alpha$, so the tamper guards inspect $\mathsf{neutral}$ rather than
pattern-matching on a variable head. We further prove one-step preservation
($\mathsf{preservation}$) and multi-step preservation
($\mathsf{preservation\_star}$) for the complete operational relation,
including the two binder-commuting $\nu$ rules, together with result
regularity ($\mathsf{typing\_regular}$) and regularity of all embedded
annotations ($\mathsf{typing\_annotations\_regular}$). These results are
collected as Theorem~\ref{thm:target-safety}.

\paragraph{The target is not strongly normalizing.}
The dynamic type permits a closed, well-typed self-application cycle. The file
\texttt{nonnormalization.v} proves \texttt{well\_typed\_non\_normalizing} and
the direct negation
\texttt{typing\_does\_not\_imply\_strong\_normalization}. Target progress and
preservation therefore establish type safety, not termination. Source CoC
strong normalization remains essential to the definition of $\nf$ used by
extraction (\S\ref{sec:extraction}).

\paragraph{Syntactic simulation.}
To state extraction correctness we use a \emph{syntactic simulation} relation
$s \sepsim t$ (Rocq: \texttt{sim}), which relates terms differing by replacing a
type argument with $\dyn$, by inserting or stripping casts, by unfolding
$\beta$/type-$\beta$ redexes, and---critically---by the rule
$\blame q \sepsim t$ for \emph{any} $t$ ($\mathsf{sim\_blame}$): a blame term
is related by $\sepsim$ to every term. Its reflexive-transitive closure is $\sepsim^{*}$
(\texttt{sim\_star}). We use $\sepsim$ only as a syntactic relation to state
extraction correctness. The development does not define a typed grammar of
program contexts or a contextual approximation theorem. Indeed, the current untyped
$\sepsim$ (which includes rules such as
$\mathsf{sim\_left\_tabs}\colon s \sepsim t \Rightarrow \Lambda K.\,s \sepsim t$)
is not a behavioral preorder, since $\Lambda K.\,s$ and $s$ are
distinguished by the context $[\cdot]\,[\dyn]$. We therefore treat every
correctness statement below as a statement about $\sepsim^{*}$. One algebraic
lemma says that a type application at a concrete type is $\sepsim$-related to
the same application at $\dyn$; no behavioral reading is attached to it.

%% ==========================================================================
\section{The Extraction}\label{sec:extraction}
%% ==========================================================================

The extraction comprises three mutually supporting pieces: a translation of
kinds, a translation of types, and a translation of terms driven by typing
derivations. All three respect definitional equality: types are normalized
before translation, and binder classification uses the reduction-stable
$\islarge$.

\paragraph{Kinds.}
A large source type (a kind) becomes a target kind by turning classifier domains
into kind arrows and dropping term domains:
\[
\extk(s) = \kstar, \quad
\extk(\Pi T.\,U) = \begin{cases}
  \extk(T) \karr \extk(U) & \text{if $T$ is a classifier} \\
  \extk(U) & \text{otherwise.}
\end{cases}
\]
Thus $\Set \to \Nat \to \Set$ becomes $\kstar \karr \kstar$: the type argument
$\Set$ contributes a $\kstar$, the term argument $\Nat$ is dropped. This is
exactly the kind a vector-like family needs.

\paragraph{Types.}
Type translation is $\extt(\Gamma, T) = \extt_L(\Gamma, \nf(T))$: normalize with
$\nf$ (justified by strong normalization), then translate the normal form with
$\extt_L$ (Figure~\ref{fig:extt}). A sort becomes $\dyn$; a variable becomes a
type variable if its binding is large, else $\dyn$; a product becomes a
$\forall$ (large domain) or an arrow (small domain); a type-operator
$\lambda$ becomes $\Lambda$ (large domain) or is erased (small domain); and an
application becomes a type application when both sides are type-level, drops a
term argument when only the head is type-level, and otherwise becomes $\dyn$.
The auxiliary $\mathsf{typeexpr}$ decides whether a term denotes a type-level
object. Because classification here consults $\islarge$ through the context and
because the argument is already normal, $\extt$ is invariant under conversion of
both the type and the context.

\begin{figure}
\small
\begin{align*}
\extt_L(\Gamma, s) &= \dyn \\
\extt_L(\Gamma, n) &= \begin{cases}
   \alpha_{\,\mathrm{idx}(\Gamma,n)} & \text{if $\islarge$ at $n$} \\
   \dyn & \text{otherwise}
 \end{cases}\\
\extt_L(\Gamma, \Pi T.\,U) &= \begin{cases}
   \forall{:}\extk(T).\,\extt_L(\Gamma{,}T, U) & \text{$T$ classifier}\\
   \extt_L(\Gamma, T) \to \extt_L(\Gamma{,}T, U) & \text{otherwise}
 \end{cases}\\
\extt_L(\Gamma, \lambda T.\,M) &= \begin{cases}
   \Lambda{:}\extk(T).\,\extt_L(\Gamma{,}T, M) & \text{$T$ classifier}\\
   \extt_L(\Gamma{,}T, M) & \text{otherwise}
 \end{cases}\\
\extt_L(\Gamma, u\ v) &= \begin{cases}
   \extt_L(\Gamma,u)\ \extt_L(\Gamma,v) & \text{$u,v$ type-level}\\
   \extt_L(\Gamma, u) & \text{$u$ type-level, $v$ not}\\
   \dyn & \text{otherwise}
 \end{cases}
\end{align*}
\caption{Type translation $\extt_L$ on normal forms. Term arguments of a
type-level head are dropped; a type-level application is kept; everything else
falls back to $\dyn$.}
\label{fig:extt}
\Description{Translation rules for source types into target types.}
\end{figure}

\paragraph{Terms.}
The term translation $\ext$ recurses on a \emph{typing derivation}
(Figure~\ref{fig:ext}). Two helpers absorb the impedance mismatch. The
\emph{coercion} $\coerce{s}{A}{B}$ is the identity when $A = B$; otherwise, if
the executable compatibility judgment $\mathsf{compat}(A,B)$ holds (decidable,
\S\ref{sec:target}), a single direct cast; and otherwise reserved internal
blame, marking a coercion for which this compatibility relation has no case:
\[
\coerce{s}{A}{B} =
  \begin{cases}
    s & A = B \\
    \cast{s}{A}{B}{\lint} & A \neq B,\ \mathsf{compat}(A,B) \\
    \blame \lint & \text{otherwise.}
  \end{cases}
\]
The \emph{placeholder} $\dyntoken = \gnd{(\lambda{:}\dyn.\,0)}{(\dyn \to \dyn)}$
is an inert, closed, well-typed inhabitant used where a source sort or product
sits in term position.

The interesting cases are \textsc{Var} and \textsc{App}. A variable whose type
is large is a \emph{type variable used as a term}---a construct with no direct
term-level image in this translation---and extracts to
$\coerce{(\blame \lint)}{\dyn}{\cdot}$:
an \emph{internal} blame. A small variable extracts to a coerced term variable.
An application whose function domain is large is a type application ($\Lambda$
elimination) and extracts to a \emph{bare} type application $e\,[A]$: its
natural target type $\extt(V{:}\Gamma, U_r)[\extt(\Gamma,v)/0]$ and the expected
result type $\extt(\Gamma, U_r[v])$ are equal only up to F$\omega$ definitional
equality $\mathsf{ty\_equiv}$ (\S\ref{sec:target}), not on the nose, so this case
relies on the target conversion rule \textsc{T-Conv} rather than a runtime
coercion (see ``extraction does not commute with substitution syntactically''
below). A small application extracts to a coerced $e\ e'$. Conversions extract to
coercions between the extracted source and target types---which, by
conversion-invariance of $\extt$, are equal, so the coercion is the identity.
Sorts and products in term position extract to $\dyntoken$.

\begin{figure}
\small
\begin{align*}
\ext(\textsc{Prop}) &= \ext(\textsc{Set}) = \dyntoken \\
\ext(\textsc{Var}\ n) &= \begin{cases}
   \coerce{(\blame \lint)}{\dyn}{\extt(\Gamma, T)} & \islarge(\Gamma, T)\\
   \coerce{x_{\mathrm{idx}(\Gamma,n)}}{\extt_{\mathrm{lookup}}(\Gamma, n)}{\extt(\Gamma, T)} & \text{otherwise}
 \end{cases}\\
\ext(\textsc{Lam}) &= \begin{cases}
   \Lambda{:}\extk(T).\,\ext(M) & \islarge(\Gamma, T)\\
   \lambda{:}\extt(\Gamma, T).\,\ext(M) & \text{otherwise}
 \end{cases}\\
\ext(\textsc{App}) &= \begin{cases}
   \ext(u)\,[\extt(\Gamma, v)] & \islarge(\Gamma, V)\\
   \coerce{(\ext(u)\ \ext(v))}{\cdot}{\cdot} & \text{otherwise}
 \end{cases}\\
\ext(\textsc{Prod}) &= \dyntoken \\
\ext(\textsc{Conv}) &= \coerce{\ext(t)}{\extt(\Gamma, U)}{\extt(\Gamma, V)}
\end{align*}
\caption{Term translation $\ext$, by recursion on the typing derivation. The
large \textsc{App} case emits a bare type application, well typed at its natural
type $\extt(V{:}\Gamma,U_r)[\extt(\Gamma,v)/0]$ and reconciled with the expected
result type $\extt(\Gamma,U_r[v])$ by the target conversion rule \textsc{T-Conv}
(the two are $\mathsf{ty\_equiv}$-equal but not syntactically equal). The elided
coercion targets in the small cases are the substituted vs.\ actual result types;
there they are equal by conversion-invariance of $\extt$, so the coercion is the
identity. The elided source types $T, V, U$ are those of the corresponding
derivation nodes.}
\label{fig:ext}
\Description{Translation rules for source typing derivations into target terms.}
\end{figure}

\paragraph{Extraction does not commute with substitution syntactically.}
The large \textsc{App} case rests on a subtle point: type extraction commutes
with the source substitution of the application rule in general only \emph{up
to} $\mathsf{ty\_equiv}$, not uniformly on the nose. Concretely, for
$u : \Pi V.\,U_r$ with
$V$ large and argument $v : V$, the two sides
\[
  \extt(V{:}\Gamma, U_r)\,[\,\extt(\Gamma, v)\,/\,0\,]
  \qquad\text{and}\qquad
  \extt(\Gamma,\, U_r[v])
\]
are $\mathsf{ty\_equiv}$-equal but syntactically distinct, for two independent
reasons. \emph{(i) Target $\beta$.} Plugging a large argument can place a
type-operator abstraction in head position, so the left side may contain a
target redex $(\Lambda \alpha{:}K.\,A)\ B$ whose contractum is what appears on the right;
these are related only by the F$\omega$ type-reduction $\mathsf{ty\_step}$.
\emph{(ii) Annotation drift under $\nf$.} Source substitution rewrites binder
annotations up to conversion, and $\extt = \extt_L \circ \nf$ collapses those to
a common normal form; the raw (pre-$\nf$) syntaxes differ. A naive
substitution-commutation lemma phrased with syntactic equality is therefore
\emph{false}. The correct statement (Rocq: \texttt{extract\_typ\_tsubst\_coc\_equiv})
replaces $=$ with $\mathsf{ty\_equiv}$, and its $\beta$ case is discharged by
$\mathsf{ty\_equiv\_beta}$. This is exactly why the extractor emits a bare $e\,[A]$
and defers to \textsc{T-Conv}: the target's F$\omega$ conversion rule
absorbs precisely this definitional equality, so no runtime cast is needed.
The same phenomenon defeats the naive \emph{kind} index of type extraction
(\texttt{extract\_kind\_L} does not commute with the substitution either, since
substituting a type for a variable can flip a domain's syntactic classifier), so
kind-regularity (\texttt{extract\_typ\_wf\_sort}) must thread the target kind
through the CoC kinding derivation rather than read it off the CoC type
syntactically.

\paragraph{Optimism, made precise.}
Two design choices motivate the name \emph{optimistic}. First, term indices of
type families are \emph{dropped}, not
dynamized: \texttt{Vec A n} extracts to the F$\omega$ type application
$\extt(\Gamma,\texttt{Vec})\,\extt(\Gamma,\texttt{A})$, not to something
mentioning $\dyn$. Second, the only sources of $\dyn$ in extracted \emph{types}
are the fallback cases in Figure~\ref{fig:extt} (a sort, a
term-level variable, or a term-headed application appearing in type position).
The examples in \S\ref{sec:overview} show both erasable indices and a
value-as-type boundary. We characterize exactly when this particular type
translation emits no $\dyn$ in Theorem~\ref{thm:dynfree}; this is a syntactic
characterization, not a semantic optimality claim.

%% ==========================================================================
\section{Metatheory}\label{sec:metatheory}
%% ==========================================================================

All results below are mechanized in Rocq and are admit-free. The audited core
set uses only the standard-library \texttt{eq\_rect\_eq}
(\S\ref{sec:mech}). We name each
Rocq theorem for cross-reference. Let $\ext(\Gamma \vdash t : T)$ denote the
term extraction of a derivation of $\hastype{\Gamma}{t}{T}$, and
$\extc(\Gamma)$ the extracted context.

\begin{theorem}[Target regularity and safety; \texttt{typing\_regular},
\texttt{typing\_annotations\_regular}, \texttt{progress},
\texttt{preservation}]
\label{thm:target-safety}
Typing in a well-formed target context produces a kind-$\kstar$ result and
well-formed annotations. If $\hastype{\emptyset}{e}{A}$, then $e$ is a value,
$e=\blame p$, or $e\reduces e'$ for some $e'$. If
$\hastype{\Delta}{e}{A}$ and $e\reduces e'$, then
$\hastype{\Delta}{e'}{A}$; the result extends to $\reducesstar$.
\end{theorem}

\begin{theorem}[Well-typed extraction; \texttt{extract\_well\_typed}]
\label{thm:welltyped-extraction}
For every derivation of $\hastype{\Gamma}{t}{T}$ with $\Gamma$ well-formed,
\[
\hastype{\extc(\Gamma)}{\ext(\Gamma \vdash t : T)}{\extt(\Gamma, T)}.
\]
\end{theorem}
The extraction of a well-typed CoC term is well typed in the kind-regular
target judgment, at the extracted type. The proof is by induction on the derivation, using
conversion-invariance of $\extt$ to discharge the coercion cases and the typing
of $\dyntoken$ and $\blame \lint$ at $\dyn$.

\begin{theorem}[Simulation; \texttt{extract\_reduces\_once}, \texttt{extract\_reduces}]
\label{thm:sim}
If $\hastype{\Gamma}{t}{T}$ and $t \reduces v$ (so, by subject reduction,
$\hastype{\Gamma}{v}{T}$), then
\[
\ext(\Gamma \vdash t : T) \;\sepsim^{*}\; \ext(\Gamma \vdash v : T).
\]
The multi-step version follows by transitivity of $\sepsim^{*}$.
\end{theorem}
Source reduction is tracked by target simulation. The proof develops the
expected substitution lemmas at the target: extraction commutes with term
substitution and with type substitution up to $\sepsim$, and $\beta$/type-$\beta$
source redexes map to the corresponding $\mathsf{sim\_beta}$/$\mathsf{sim\_tbeta}$
steps. Weakening lemmas (\texttt{extract\_weaken}$\ast$) handle the shift in
de~Bruijn indices between the two namespaces.

This theorem is deliberately \emph{syntactic}. Because $\sepsim$ contains the
one-sided rule $\blame q \sepsim t$, an extraction that has already fallen to
reserved internal blame on the left simulates the reduct trivially: the relation
records that the target has entered an internal-residue case rather than
asserting behavioral preservation. A behavioral (contextual-approximation)
statement would need a typed reformulation of $\sepsim$ and a CIU/logical-relations
adequacy argument; that bridge is open (\S\ref{sec:limitations}).

\begin{theorem}[Derivation independence; \texttt{extract\_deriv\_indep}]
\label{thm:indep}
Any two derivations $H_1, H_2$ of the same judgment $\hastype{\Gamma}{t}{T}$
satisfy $\ext(\Gamma \vdash_{H_1} t : T) \sepsim^{*} \ext(\Gamma \vdash_{H_2} t :
T)$.
\end{theorem}
Because $\ext$ recurses on derivations, this licenses ``the extraction'' as
shorthand only up to the syntactic simulation relation: different derivations
are $\sepsim^{*}$-related. It
follows from a conversion-strengthening lemma
(\texttt{extract\_deriv\_indep\_conv}) applied to reflexivity.

\begin{theorem}[External-label non-generation; \texttt{extraction\_blame\_free}]
\label{thm:blamefree}
For every derivation of $\hastype{\Gamma}{t}{T}$ and every label $p$ with
$\mathrm{lbl\_id}(p) \ge 3$,
\[
\neg\,\bigl(\ext(\Gamma \vdash t : T) \reducesstar \blame p\bigr).
\]
\end{theorem}
An extracted term reduced in isolation never reaches an external label. The
proof combines a safety lemma (\texttt{extracted\_safe}: every cast the
extraction introduces uses the internal label $\lint$, hence is safe for any
external $p$) with Theorem~\ref{thm:blame}. The reserved identifiers are
excluded: 0 for $\nu$ tampering, 1 for $\mathsf{is\_gnd}$ tampering, and 2 for
the extraction/coercion failure label $\lint$. Because the source calculus has
no external cast labels and the theorem quantifies over no linking context,
this is an external-label non-generation theorem, not an interoperability
guarantee for arbitrary programmer boundaries.

\begin{theorem}[Syntactic instantiation; \texttt{extraction\_instantiation\_sim}]
\label{thm:instantiation}
Let $H$ derive $\hastype{\Gamma}{M}{\Pi \srt.\,A}$ for a sort $\srt$, and let
$C$ and $A$ be strongly normalizing (the witnesses used to form the extracted
annotations). Then for every target type $B$ and label $p$, instantiating the
extracted term at $\extt(\Gamma, C)$ is related by $\sepsim$ to instantiating
it at $\dyn$:
\[
\begin{array}{c}
\cast{(\ext(H)\,[\extt(\Gamma,C)])}{\extt(\srt{:}\Gamma,\,A)[\extt(\Gamma,C)/0]}{B}{p}\\[4pt]
\sepsim\;\;
\cast{(\ext(H)\,[\dyn])}{\extt(\srt{:}\Gamma,\,A)[\dyn/0]}{B}{p}.
\end{array}
\]
\end{theorem}
This is an extraction-level instance of
\texttt{cast\_instantiation\_sim}: the concrete instantiation is
$\sepsim$-related to the $\dyn$ instantiation. Its premises require only the
normalization witnesses used to form the extracted annotations, not source
typing of $C$ or contextual compatibility. It is therefore intentionally not
called the Jack-of-All-Trades theorem.

\begin{theorem}[Optimism / $\dyn$-freedom; \texttt{extract\_typ\_dyn\_free\_iff}]
\label{thm:dynfree}
For the recursively defined source predicate $\mathsf{fo}$ (Rocq:
\texttt{fo\_typ\_L})---which rejects
sorts and term-level variables in type position, requires a type-level head for
applications, and recursively checks every component preserved by type
extraction---the extracted type avoids $\dyn$ \emph{if and only if} the normal
form satisfies $\mathsf{fo}$:
\[
\mathsf{typ\_dyn\_free}(\extt(\Gamma, T))
\;\Longleftrightarrow\;
\mathsf{fo}(\Gamma, \nf(T)).
\]
\end{theorem}
This is an exact characterization relative to the recursively defined
predicate $\mathsf{fo}$; it is not a semantic maximality theorem. Note that
$\mathsf{fo}$ includes many indexed-family uses such as \texttt{Vec A n},
because the term index \texttt{n} is erased from the target type
($\extt(\Gamma,\texttt{Vec A n}) = \extt(\Gamma,\texttt{Vec})\,\extt(\Gamma,\texttt{A})$);
it excludes term-headed type computation such as \texttt{El c}, where a term
determines a type.
For source types satisfying the predicate, the target type is $\dyn$-free. For
source types outside it, the converse proves that the extracted type contains
$\dyn$.

\medskip
Taken together: extraction produces well-typed target terms
(\ref{thm:welltyped-extraction}), tracks computation up to $\sepsim^{*}$
(\ref{thm:sim}), is derivation-related up to syntactic simulation
(\ref{thm:indep}), does not generate external labels under isolated reduction
(\ref{thm:blamefree}), relates concrete and $\dyn$ instantiations syntactically
(\ref{thm:instantiation}), and is $\dyn$-free exactly for the defined source-type predicate
(\ref{thm:dynfree}).

%% ==========================================================================
\section{Mechanization}\label{sec:mech}
%% ==========================================================================

The development is written in Rocq~9.0 and builds with \texttt{dune}, and is
\emph{entirely admit-free}: the target calculus (\texttt{BlameFOmega}),
well-typed extraction (\texttt{extract\_well\_typed}), the source-reduction
simulation (\texttt{extract\_reduces\_once}/\texttt{extract\_reduces},
Theorem~\ref{thm:sim}), and the derivation-independence and external-label
non-generation theorems all carry complete machine-checked proofs. The extraction
weakening/substitution commutation lemmas the simulation needs---%
\texttt{term\_tlift\_extract\_sim}, \texttt{term\_tsubst\_extract\_sim},
\texttt{extract\_weaken1}, \texttt{extract\_subst\_sim\_gen},
\texttt{extract\_tsubst\_gen}---are the ``$\nf$ commuting with
\texttt{lift}/\texttt{subst}'' inductions, parallel to
\texttt{extract\_typ\_L\_swap}'s treatment of reduction.

The two lemmas discharging the large-\textsc{App} \textsc{T-Conv} premises are
both proved. Kind regularity, \texttt{extract\_typ\_wf\_sort} (type extraction is
well kinded at $\kstar$), is axiom-free via a stronger induction
(\texttt{extract\_typ\_L\_wf\_kind}) that threads the target kind through the CoC
kinding derivation restricted to genuinely type-level subterms
(\texttt{type\_expr}), which is what makes the otherwise-false naive kind index
correct. The conversion premise, \texttt{extract\_typ\_tsubst\_coc\_equiv}
(type extraction commutes with the application rule's source type substitution
\emph{up to} $\mathsf{ty\_equiv}$; \S\ref{sec:extraction}), factors into a raw
structural substitution-commutation lemma
(\texttt{extract\_typ\_L\_large\_subst\_raw}, plain structural induction---%
\texttt{extract\_typ\_L} never fires a redex) and reduction-invariance of
\texttt{extract\_typ\_L} up to $\mathsf{ty\_equiv}$
(\texttt{extract\_typ\_L\_reduces\_nf\_equiv}, a single-step lemma chained over
$W \to^{*} \nf\,W$); the single-step $\beta$ case defers to the raw lemma via
$\mathsf{ty\_equiv\_beta}$. The repository includes
\texttt{Extraction/Assumptions.v}, which audits a listed set of headline
theorems; the only assumption that appears in that set is the standard library's
\texttt{eq\_rect\_eq} instance introduced by dependent \texttt{destruct} (the
REPL name-recovery lemmas additionally use the standard primitive
string/integer axioms, since they manipulate variable names; the core metatheory
does not). A source scan separately rejects proof-hole and axiom-declaration
commands, and both checks run under \texttt{dune test}. The original
Coq-in-Coq base itself remains zero-admit.

\paragraph{Type-valued derivations.}
The single most consequential engineering decision is that $\ext$ recurses on a
typing \emph{derivation}, and a derivation carries computational content only if
the typing judgment lives in \texttt{Type}, not \texttt{Prop}: Rocq forbids
eliminating a \texttt{Prop} inductive into a \texttt{Type}-returning function.
We therefore moved \texttt{has\_type} and \texttt{well\_formed} to \texttt{Type},
which cascaded through the whole reduction/conversion layer: \texttt{reduces},
\texttt{reduces\_once}, \texttt{convertible}, and the parallel-reduction
relations all became \texttt{Type}-valued, with hand-rolled inductives replacing
the standard-library \texttt{clos\_*} closures. Strong normalization, which is
naturally a \texttt{Prop}, is bridged to the \texttt{Type} world by a wrapper
(\texttt{reduces\_once\_prop}). Numerous \texttt{exists}/$\vee$ were changed to
$\Sigma$-types/$+$ across roughly twenty files, all while retaining the original
proofs' structure.

\paragraph{Normalization and decidability inside the function.}
The extraction calls $\nf$ (compute a normal form, total by strong
normalization) and $\islarge$-decision (\texttt{is\_large\_dec}, the type-checking
decision procedure) to classify context bindings in a conversion-invariant way.
These are well-defined total functions, so $\ext$ is a total mathematical
function on derivations; but they make $\ext$ a \emph{metatheoretic} object
rather than a term that reduces under \texttt{reflexivity}
(\S\ref{sec:limitations}). Conversion-invariance is itself a theorem:
$\islarge$ is stable under reduction (\texttt{is\_large\_stable}), and $\extt$ is
invariant under reduction of both its type argument and its context
(\texttt{extract\_typ\_L\_swap}, \texttt{extract\_ctx\_swap}).

The OCaml extraction step warns that it accesses opaque proof bodies used by
computational, \texttt{Type}-valued decision procedures. This does not change
the kernel-checked theorem statements, but it means the generated REPL is not a
separately verified compiler with a small transparent algorithmic trusted base.

\paragraph{Proof effort.}
The development is about 27{,}500 lines of Rocq. The adapted Coq-in-Coq
\texttt{CoC} layer is roughly 10{,}900 lines; the \texttt{BlameFOmega} target
(syntax, semantics, kinding/typing, subtyping, safety, label-safety theorems,
confluence, and syntactic simulation) is a self-contained mechanization of
roughly 8{,}300 lines; and the \texttt{Extraction} layer---split into
translation, typing, substitution-simulation, reduction-simulation, and
related modules---is another roughly 8{,}300 lines, dominated by the
substitution/weakening infrastructure underlying Theorem~\ref{thm:sim}. The
unverified OCaml shell (REPL loop, Menhir parser, and string layout) is under
600 lines.

%% ==========================================================================
\section{A REPL with Verified Name Recovery}\label{sec:repl}
%% ==========================================================================

To make the extraction inspectable we extract the type checker and the
translation to OCaml (via Rocq's extraction) and wrap them in a REPL that reads
CoC surface syntax and prints inferred types and F$\omega$ + blame extractions
in Rocq-like concrete syntax (\texttt{fun} and \texttt{forall} with grouped
binders, \texttt{cast<A => B>\^{}$\ell$(e)} for casts, and $\dyn$ for the
dynamic type). The paper's output comments are transcribed from the
checked-in golden files (which \texttt{dune test} replays), with line breaks
and indentation added for readability. The main commands
are \texttt{Axiom}, \texttt{Definition}, \texttt{Inductive}, \texttt{Check},
\texttt{Infer}, \texttt{Compute}, and \texttt{Extract}, plus \texttt{Print
Axioms} to list the current context; \texttt{Check e.} without a type ascription
behaves like \texttt{Infer}.
\texttt{Definition} is handled in the unverified shell by \emph{delta-unfolding}:
before a command reaches the verified checker, each reference to a definition is
replaced by its body (a capture-avoiding substitution), so the checker only ever
sees a definition-free term and definitional transparency reduces to $\beta$.
\S\ref{sec:examples} lists complete input files.

\paragraph{Inductive types by Boehm--Berarducci encoding.}
The shell also offers an \texttt{Inductive} command that is pure sugar for the
impredicative Boehm--Berarducci encoding~\cite{bohm1985automatic}
(equivalently, Pfenning and Paulin-Mohring's encoding of inductively defined
types in the Calculus of Constructions~\cite{pfenning1990inductively}). It
generates ordinary \texttt{Definition}s---the type former, the constructors, and
a recursor $T\_\mathit{rec}$---so, being definitions rather than axioms, the
data \emph{computes}: a companion \texttt{Compute} command reduces a term to
normal form with the verified normalizer (e.g.\ \texttt{Compute}-ing $2+1$ over
an \texttt{Inductive Nat} returns the Church numeral for $3$). The encoding is
\emph{indexed}: for an arity $I_1 \to \cdots \to I_k \to s$ the generated motive
$X : I_1 \to \cdots \to I_k \to s$ is parameterized by the indices, each
constructor may target $T\,\overline{p}\,\overline{t}$ at arbitrary index
values, and a recursive argument $T\,\overline{p}\,\overline{u}$ is folded to
$X\,\overline{u}$. This suffices to define \texttt{Vec}, \texttt{Fin}, Leibniz
equality (whose $Eq\_\mathit{rec}$ is exactly transport), the \texttt{Set}-indexed
format family \texttt{Fmt}, session channels, and inductive predicates such as
$\le$ and reflexive--transitive closure---all as computing $\lambda$-terms.

\paragraph{A typeability/computability tension.}
Two boundaries here are intrinsic to \emph{any} impredicative encoding, not to
our implementation. First, the generated recursor's motive depends on the
\emph{indices} but not on the constructed \emph{value}: it is a catamorphic
recursor, not the dependent induction principle, and induction is provably
\emph{not derivable} for impredicative encodings in second-order dependent type
theory~\cite{geuvers2001induction}. So large elimination (a motive over a
non-indexed value, as in \texttt{vreplicate}) and dependent projections (a
$\Sigma$-type's second projection) stay out of reach. Second, one-step
\emph{inversion}---case analysis exposing an immediate subterm, as in
\texttt{Fin}'s successor case or a heterogeneous-list head---is not a fold and is
likewise not derivable, so such principles are kept axiomatic. Scott and
Parigot encodings do give constant-time inversion and predecessors, but their
types are \emph{recursive} and hence not typeable in a strongly-normalizing pure
type system that lacks $\mu$-types~\cite{parigot1992recursive}; Mendler-style
encodings~\cite{mendler1991inductive} tame recursion by quantifying over the
recursion carrier and accommodate wider (mixed-variance) functors, but share the
same dependency ceiling. Recovering \emph{dependent} elimination for
lambda-encoded data with computation requires self
types~\cite{fu2014self}, an extension beyond CoC. Accordingly, the example files
adapted to \texttt{Inductive} exercise exactly the fragment the encoding
supports; the handful of operations needing value-dependent elimination or
inversion retain a clearly-marked axiom.

\paragraph{Verified de~Bruijn-to-named printing.}
Metatheory uses de~Bruijn indices, but readable output needs names. Naively
re-inventing names (\texttt{x0}, \texttt{x1}, \dots) is unreadable and, if done
by hand in the printer, unverified and capture-prone. We instead verify the
conversion. On the source side, \texttt{expression\_of\_term\_with\_hints}
converts a de~Bruijn term to a named expression, reusing the programmer's
original binder names as hints; on the target side,
\texttt{fterm\_expression\_of} does the same for F$\omega$ + blame across its two
namespaces. Its source-binder hint stream is separate from names generated
inside type annotations, so an annotation cannot consume the next term-binder
name. Both conversions are proved to produce output
\emph{equivalent} to the de~Bruijn input (an inductive equivalence relation
carried in the function's result type), and the shared name chooser
\texttt{pick\_name} accepts a hint only when it is fresh for the current context,
so recovered names are \emph{capture-free by construction}---the freshness proof
is part of \texttt{pick\_name}'s type. We further prove the hinted conversion is
alpha-equivalent to the fresh-name conversion (\texttt{..\_alpha}), and that a
named expression determines its de~Bruijn preimage uniquely
(\texttt{..\_unique}). The extracted
\texttt{fterm\_expression\_of\_checked} wrapper decides namespace uniqueness and
de~Bruijn scope before invoking the proof-carrying conversion; it returns
\texttt{None} if either check fails. Only final string layout remains in the
unverified OCaml shell.

%% ==========================================================================
\section{Worked Examples}\label{sec:examples}
%% ==========================================================================

Table~\ref{tab:examples} summarizes the eighteen showcase files shipped with
the development (three additional files test naming and error paths). Each is
a self-contained command batch processed by the REPL. Wherever the datatype is
expressible by the indexed Boehm--Berarducci encoding, the file declares it with
the \texttt{Inductive} command (\S\ref{sec:repl}), so its type former,
constructors, and recursor are computing definitions---this covers the
non-indexed types, the length-indexed \texttt{Vec}, Leibniz equality, session
channels, sorted lists over an inductive $\le$, and more. The remaining
inductive kit is axiomatized exactly where the encoding cannot reach: the
value-as-type boundaries (\texttt{El}, \texttt{eval}) that need large
elimination, and one-step inversion principles (\texttt{Fin}'s
\texttt{fin\_case}, heterogeneous-list heads) that are not folds. Every other
operation, from \texttt{vnth} through matrix multiplication and \texttt{sprintf}
to the proof steps of Newman's lemma, is an ordinary \texttt{Definition}.

\begin{table}
\small
\caption{The eighteen showcased examples. ``$\dyn$/blame'' marks examples that
exhibit the value-as-type boundary under our extraction.
``\texttt{Obj.magic}'' marks examples whose idiomatic-CIC port extracts with
unchecked casts or \texttt{Obj.t}-typed interfaces under Rocq~9.0's stock
OCaml extraction (see the experiment below).}
\label{tab:examples}
\Description{A table of eighteen example files, the dependent programming idiom each demonstrates, whether it exhibits the dynamic-type boundary, and whether stock extraction of its CIC port is unsafe.}
\begin{tabular}{@{}llcc@{}}
\toprule
File & What it demonstrates & $\dyn$/blame & \texttt{Obj.magic} \\
\midrule
\texttt{newman.v}   & Newman's lemma over axiomatized inductive predicates & & \\
\texttt{vectors.v}  & length-indexed vector signatures and clients & & \\
\texttt{fin.v}      & bounds-checked indexing; length index erased & \checkmark & \checkmark \\
\texttt{equality.v} & equality indices erased from types; proof terms remain & & \\
\texttt{sigma.v}    & dependent pairs; existential over a size & & \\
\texttt{printf.v}   & type-safe \texttt{printf}; dependent arity & & \checkmark \\
\texttt{matrix.v}   & dimension-indexed matrices; shape safety & \checkmark & \\
\texttt{units.v}    & units of measure; dimension index erased & \checkmark & \\
\texttt{avl.v}      & height-indexed balanced trees & & \\
\texttt{session.v}  & session-typed protocols; typestate & & \\
\texttt{stlc.v}     & intrinsically typed interpreter signature & & \checkmark \\
\texttt{universe.v} & Tarski universe; a hand-built dynamic type & \checkmark & \checkmark \\
\texttt{functor.v}  & higher-kinded: functors, tagless-final, monads & & \checkmark \\
\texttt{tagless.v}  & computed return types via type-level codes; $\dyn$-free & & \checkmark \\
\texttt{lists.v}    & polymorphic list library; fully static baseline & & \\
\texttt{scoped.v}   & intrinsically well-scoped terms; capture-avoiding renaming & \checkmark & \\
\texttt{hlist.v}    & heterogeneous lists indexed by a type-level list of types & \checkmark & \checkmark \\
\texttt{ordered.v}  & provably-sorted lists over the \texttt{Le} inductive predicate & \checkmark & \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Baseline: stock extraction of the same idioms is unsafe, not just
untyped.}
To calibrate what the blame target buys, we ported each showcased file to
idiomatic CIC and ran Rocq~9.0's stock OCaml extractor on the ports
(\texttt{paper/stock-extraction/} in the artifact reproduces the experiment).
The ports are the programs a Rocq user would actually write: the REPL's
\texttt{Inductive} encodings become real \texttt{Inductive} declarations, and
the kit that the bare PTS had to axiomatize becomes genuine
definitions---one-step inversions (\texttt{fin0}, \texttt{fin\_case},
heterogeneous-list projections) become dependent \texttt{match}es, value-to-type
decoders (\texttt{El}) become \texttt{Fixpoint}s by large elimination,
\texttt{stlc.v}'s axiomatized evaluator is implemented, and \texttt{tagless.v}
is ported as the \texttt{match}-computed-return-type CIC program it emulates.
The result (Table~\ref{tab:examples}, last column): seven of the eighteen
extract with \texttt{Obj.magic} casts in their bodies (\texttt{fin},
\texttt{hlist}, \texttt{printf}, \texttt{functor}, \texttt{stlc},
\texttt{tagless}) or with \texttt{Obj.t} in their interfaces
(\texttt{universe}, whose decoded payload is exposed at type \texttt{Obj.t},
pushing the unchecked cast onto every client).  The eleven clean files are
exactly those whose dependency erases into ML polymorphism or \texttt{Prop}.
The two columns of Table~\ref{tab:examples} measure different boundaries and
disagree in both directions: \texttt{matrix.v} is clean under stock
extraction but checked under $\dyn$/blame only because the file deliberately
extracts the type family itself as data, while \texttt{tagless.v} is
$\dyn$-free under our extraction (its computed return type is a genuine
F$\omega$ type-level application, \S\ref{sec:examples}) yet its CIC port is
among the heaviest \texttt{Obj.magic} users under stock extraction.

The unsafe seven are not merely inelegant.  The erased index is what made the
source safe, so the extracted module admits well-typed OCaml calls that
violate it, and \texttt{Obj.magic} turns the violation into undefined
behavior.  Concretely, the port of \texttt{fin.v} extracts the empty-index
eliminator and lookup as
\begin{lstlisting}[language={}]
let fin0 _ = Obj.magic Tt
let rec nth _ = function
  | Nil -> fin0
  | Cons (m, a, v) -> (fun i -> fin_case m a (nth m v) i)
\end{lstlisting}
with type \texttt{nat -> 'a1 t -> fin -> 'a1}: the bound linking the vector
and the index is gone from the OCaml type, so the out-of-bounds call is
well typed.  The client
\begin{lstlisting}[language={}]
let f : int -> int = nth O Nil (F1 O) in f 3
\end{lstlisting}
compiles without warning and dies with SIGSEGV: \texttt{nth} reaches
\texttt{fin0}, whose \texttt{Obj.magic Tt} materializes a unit value at type
\texttt{int -> int}, and applying it crashes the process.  No OCaml exception
is raised, so the failure cannot be caught, and nothing identifies the
violated boundary.  In the blame target the corresponding boundary is a typed
cast: a well-typed client linked against an extracted term forms a well-typed
target program, and target progress and preservation
(\S\ref{sec:metatheory}) confine every failure to an explicit, observable
$\blame{\ell}$ term whose label names the boundary.  Blame is a defined
outcome of the operational semantics---runtime-safe in the sense of the
progress theorem---where \texttt{Obj.magic} is a promise nobody checked.
(Which label can arise where is a separate question; our external-label
theorem covers isolated extracted terms only, \S\ref{sec:limitations}.)

We highlight four that best illustrate the design.

\paragraph{The computed return type that needs no $\dyn$ (\texttt{tagless.v}).}
The CIC program one most expects to force the dynamic type is an
intrinsically typed interpreter whose \emph{result type is computed from a
run-time tag}:
\begin{lstlisting}
Fixpoint eval {t} (e : expr t)
  : match t with TNat => nat | TBool => bool end := ...
\end{lstlisting}
\texttt{tagless.v} expresses this program in bare CoC---no inductives, no
universes, no large elimination---by Church-encoding the object types
\emph{one level up}, as type-level booleans of kind $\kstar \karr \kstar
\karr \kstar$ (\texttt{TNat} $= \lambda a\,b{:}\Set.\,a$), so that CIC's
large elimination becomes type-level \emph{application}
$t\;\mathrm{CNat}\;\mathrm{CBool}$, discharged by ordinary
$\beta$-conversion. The entire file---Church naturals and booleans, the
indexed expression family, and \texttt{eval}---extracts \emph{without a
single $\dyn$, cast, or blame}:
\[
\texttt{eval} : \forall t{:}\kstar{\karr}\kstar{\karr}\kstar.\;\;
\mathrm{expr}\;t \to t\;\mathrm{CNat}\;\mathrm{CBool},
\]
with the motive surviving as a genuine F$\omega$ type-level function. This
is the sharpest illustration of what the higher-kinded target buys: the
example lands squarely in Theorem~\ref{thm:dynfree}'s $\dyn$-free fragment.
By contrast, the term-level-code presentation (\texttt{stlc.v}'s
\texttt{El}, next) collapses every computed type into one \emph{rigid}
abstract target type---also $\dyn$-free, but only because the bare-PTS
source cannot even write the program that consumes the computed result.
Forcing $\dyn$ as a computed \emph{return} type requires large elimination
in the source (\S\ref{sec:limitations}).

\paragraph{An intrinsically typed interpreter signature (\texttt{stlc.v}).}
Object types \texttt{Ty}, contexts \texttt{Ctx}, and well-typed variables
\texttt{Var G t} index the term family \texttt{Tm G t}, so only well-typed object
programs inhabit the interface. The file axiomatizes \texttt{El}, \texttt{Env},
and \texttt{eval}; it does not implement or run an evaluator. The identity term
and assumed interpreter application extract with indices removed:
\begin{lstlisting}
Extract fun (t : Ty) => tlam emp t t (tvar (snoc emp t) t (vz emp t)).
(* ==>  fun (t : Ty) => tlam emp t t (tvar (snoc emp t) t (vz emp t)) *)

Extract fun (G : Ctx) (t : Ty) (e : Tm G t) (env : Env G) => eval G t e env.
(* ==>  fun (G : Ctx) (t : Ty) (e : Tm) (env : Env) => eval G t e env *)
\end{lstlisting}
\texttt{Tm G t}, \texttt{Env G}, and \texttt{El t} lose their indices in the
target signature. Since \texttt{eval} is an object-language axiom, the residual
application is symbolic and exercises no evaluator reduction.

\paragraph{Units of measure (\texttt{units.v}).}
A quantity \texttt{Qty d} carries a dimension \texttt{d}, a formal expression
over a group signature; multiplication multiplies dimensions and addition
demands equal ones, so ``metres $+$ seconds'' is a type error. The operations
are definitions over the family's constructor/eliminator kit
(\texttt{qty}/\texttt{qty\_rec}) with a placeholder numeric carrier:
\begin{lstlisting}
Definition speed (len time : Dim) (d : Qty len) (t : Qty time)
  : Qty (dmul len (dinv time)) :=
  qmul len (dinv time) d (qinv time t).
\end{lstlisting}
Extraction expands the definitions and erases the dimension index from every
target type: each quantity appears at the plain type \texttt{Qty}. And the
family \texttt{Qty} itself, viewed as data (\texttt{fun d => Qty d}), hits the
boundary---emitting $\dyn$ and an internal blame---because it maps a value to
a type.

\paragraph{A hand-built dynamic type (\texttt{universe.v}).}
This example meets the target head on. A universe \texttt{U} of type codes
with a decoder \texttt{El : U -> Set} lets one build \texttt{Dynamic}, a
dependent pair of a code and a value it decodes to---literally a dynamically
typed value expressed inside CoC. Codes are ordinary data and extract
cleanly; the decoder used as a function to types is the irreducible residue,
extracting to internal blame through $\dyn$ (\S\ref{sec:overview}). The file
also extracts a bare type as \emph{data} (\texttt{Extract U -> U.}), which
produces the extraction's inert placeholder
\texttt{gnd(fun (\_ : Dyn) => \_ : Dyn -> Dyn)}---the one way a
\emph{non-blame} value of type $\dyn$ arises in extracted code. It pinpoints
where source and target part ways and connects the source-level notion of a
``dynamic value'' to the target's $\dyn$.

%% ==========================================================================
\section{Discussion and Limitations}\label{sec:limitations}
%% ==========================================================================

We are deliberately precise about what is and is not established.

\paragraph{What this paper does not claim.}
This development is not a replacement for Rocq's production extractor, nor is it
a proof-erasing compiler.  The current translation treats proof terms as ordinary
small terms; what it erases are many dependent indices from target types.  The
target is an explicit cast calculus; we do not define a gradual surface
language, cast insertion, a precision relation, or a gradual guarantee. The
simulation relation used for extraction correctness is syntactic, and no typed
contextual approximation theorem is claimed. These restrictions are
deliberate.  They isolate the central question studied here: how much of a CoC
program's polymorphic skeleton can be preserved in an F$\omega$-style target, and
where must the remaining dependency become dynamic?

\paragraph{Internal blame is reachable by design.}
Theorem~\ref{thm:blamefree} excludes external labels ($\ge 3$) only. The
extraction-failure label $\lint$ (identifier $2$) is reachable: a term that exercises a
value-as-type at run time (\S\ref{sec:overview}, \texttt{universe.v}) reduces to
$\blame \lint$. This is intentional: it marks a construct for which this
translation has no target coercion. The syntactic relation can continue a proof through this point using
$\mathsf{sim\_blame}$; that rule has no contextual or behavioral soundness claim.

\paragraph{The extraction is metatheoretic, not a compiler.}
$\ext$ invokes $\nf$ (normalization) and \texttt{is\_large\_dec} (type-checking)
to make conversion-invariant decisions. It is a total function on derivations
and everything we prove about it holds, but it does not reduce under
\texttt{reflexivity} and is not intended as an efficient back end. Making a
computable, conversion-invariant extraction---perhaps by carrying classification
evidence in a bidirectional elaboration rather than recomputing it---is future
work.

Rocq extraction of the REPL currently accesses opaque bodies of computational
proofs, producing an explicit warning. The logical development remains
kernel-checked, but the executable path is not a separately verified compiler
with a transparent algorithmic trusted base.

\paragraph{Simulation is syntactic, not contextual approximation.}
Our correctness statements (Theorems~\ref{thm:sim}, \ref{thm:instantiation}) are in terms
of the syntactic simulation $\sepsim$. The development does not define typed
program contexts or export a contextual approximation claim. A behavioral
reading of the current untyped $\sepsim$ is false: the rule $\mathsf{sim\_left\_tabs}$
relates $\Lambda K.\,s$ to $s$, yet the context $[\cdot]\,[\dyn]$ distinguishes
them (one type-$\beta$-reduces, the other is stuck). A sound bridge would need a
\emph{new typed relation}, an explicit observation policy including blame,
typed program contexts, and a CIU- or logical-relations adequacy argument.
Progress, preservation, and determinism are available infrastructure, but they
do not make the existing relation context compatible.

\paragraph{Compatibility is syntactic.}
Although target typing is kind regular, $\mathsf{compat}$ itself is a
context-free executable relation and is not closed under
$\mathsf{defeq}$. Identity compatibility admits identical non-normal
annotations because the ID step does not inspect them. Operationally this
means \emph{blame is not stable under definitional equality}: a value
injected at a neutral tag and projected at a syntactically distinct but
$\mathsf{ty\_equiv}$-equal neutral type takes the CONFLICT step and blames,
even though both annotations are well kinded and definitionally equal. The
artifact does not prove a coherence theorem for choosing compatible
representatives modulo definitional equality; non-identity clients (including
the extraction, which normalizes every annotation it constructs) normalize
annotations first.

\paragraph{Scoped $\nu$ deviates semantically from Blame for All.}
Blame for All eliminates polymorphic instantiation into a global store of
type names, so a $\dyn$ value tagged with a sealed name flows freely and
blame arises only at a later incompatible projection, attributed to that
projection's label. Our $\nu$ is a scoped binder with no store: when a
sealed value would escape its $\nu$-scope inside a $\dyn$ tag, the
$\nu$-elimination rule blames the reserved label~$0$ instead
(\S\ref{sec:target}). Consequently a program as innocent as
$(\Lambda\alpha{:}\kstar.\,\lambda x{:}\alpha.\,\cast{x}{\alpha}{\dyn}{\ell})\,[B]\ v$
reduces to $\blame 0$: injecting a sealed value into $\dyn$ and returning it
is dynamically rejected here, although Blame for All accepts it. This is a
deliberate trade---it keeps typing store-free, values simple, and progress
easy---but it means label~$0$ is reachable from non-tampering code, and the
scoped design enforces a strictly stricter dynamic discipline than the
store-based original.

\paragraph{$\dyn$-freedom is syntactically characterized.}
Theorem~\ref{thm:dynfree} gives an exact iff characterization for the current
extractor: a source type's normal form satisfies $\mathsf{fo}$ exactly when its
translation is $\dyn$-free. This does not establish semantic maximality. A
translation that accepts additional source forms without $\dyn$ would have to
change the target representation or the extraction itself.

\paragraph{Inductive families are axiomatized.}
The bare PTS has no inductive types, so our examples model families and their
eliminators as \texttt{Axiom}s. This is standard for a PTS presentation and does
not affect the metatheory (axioms are just context bindings), but extending the
source to the Calculus of Inductive Constructions---with native inductives and
their large eliminations---would let programs that \emph{compute} types from data
appear, stressing the value-as-type boundary in richer ways. The quintessential
case is an interpreter typed
\texttt{match t with TNat => nat | TBool => bool}: its bare-CoC encodings
either stay fully static (\texttt{tagless.v}, type-level codes) or collapse to
a rigid abstract type (\texttt{stlc.v}'s \texttt{El}); only a source with
large elimination produces the surviving term-headed type computation that
lands in the translation's $\dyn$ fallback as a computed \emph{return} type.
In today's examples, $\dyn$ in extracted types comes from types-as-data---%
functions whose result \emph{sort} is \Set---not from computed result types.

%% ==========================================================================
\section{Related Work}\label{sec:related}
%% ==========================================================================

\paragraph{From CoC proofs to F$\omega$ programs.}
The closest prior work is Paulin-Mohring's extraction of
F$\omega$ programs from proofs in the Calculus of Constructions
\cite{paulinmohring1989fomega}.  Paulin-Mohring uses realizability and a
distinction between informative and non-informative propositions to remove the
logical part of a development; the extracted programs are non-dependent
polymorphic terms.  We build on the same insight that a substantial CoC
fragment has an F$\omega$ skeleton.  Our contribution is different: we give a
Rocq mechanization whose target is an explicit blame calculus, not pure
F$\omega$, and we use $\dyn$ and internal blame to make the non-F$\omega$
residue explicit.  Thus our extraction is optimistic and total for the
formalized CoC core, while Paulin-Mohring's extraction is a realizability-based
program extraction with proof/logical erasure.

\paragraph{Rocq extraction and verified extraction.}
Rocq's production extraction emits ML-family code---currently OCaml, Haskell, or
Scheme---and erases dependent/proof structure that those target languages cannot
type directly \cite{letouzey2003extraction,letouzey2004these}; erasure is guided
by sort information much as our large/small split is.  Forster, Sozeau, and
Tabareau verify a modern extraction pipeline from Coq to
OCaml~\cite{forster2024verified}.  Because OCaml lacks a formal semantics, their
verified core targets Malfunction, a formal specification of the OCaml
compiler's Lambda intermediate language.  The released plugin produces
\emph{untyped} Malfunction programs together with typed interface files, and the
main guarantees are operational correctness and safe interoperability for
first-order data, not well-typedness of a residual target term.  Our theorem is
therefore orthogonal: we do not target production OCaml, but we do prove that
the extracted residual term is well typed in an explicit polymorphic blame
calculus.  The two results are strong in different dimensions: theirs gives
operational correctness and safe interoperability for a production language;
ours gives well-typedness in a rich polymorphic target at the cost of
practicality.

\paragraph{Verified erasure and verified compilation.}
MetaCoq/MetaRocq verifies type checking and type/proof erasure for Coq's kernel,
including erasure to an untyped lambda calculus
\cite{sozeau2020metacoq,sozeau2020coqcoqcorrect}.
CertiCoq \cite{anand2017certicoq} verifies a compiler from Gallina toward
lower-level executable code, and \OE{}uf \cite{mullen2018oeuf} minimizes the
extraction TCB through an untyped intermediate language.  Erasure in dependently
typed \emph{programming} languages, such as Idris'
quantitative/erasure analysis \cite{brady2013idris,tejiscak2020erasure} and
erasure in pure type systems \cite{mishra2008erasure}, removes computationally
irrelevant arguments while staying within a dependent target.  These systems aim
to produce executable code with small trusted bases.  We instead isolate a typed
intermediate-language question: after dependent information is removed from
types, what polymorphic structure can still be expressed statically, and how
should the remaining dependency be represented dynamically?

\paragraph{Gradual typing, blame, and polymorphism.}
Gradual typing \cite{siek2006gradual,siek2015refined} interleaves static and
dynamic checking; blame \cite{findler2002contracts,wadler2009blame} attributes
run-time cast failures to boundaries. Ahmed et al.'s \emph{Blame for All}
\cite{ahmed2011blame} adds polymorphism, with the Jack-of-All-Trades principle
and dynamic sealing. We adapt its $\nu$-style sealing and
blame-safety structure to an F$\omega$-style target.  Two caveats are important.
First, the published proof of the Jack-of-All-Trades Principle is invalid, as
documented by the authors' erratum~\cite{ahmed2011blameerratum}; our artifact
proves only the separately named syntactic instantiation lemma. Second,
where Blame for All's ground predicate for System~F is restricted to a bare
type-variable tag, we generalize to \emph{neutral} F$\omega$ proper types
($N ::= \alpha \mid N\,A$, \S\ref{sec:target}), so a type-family application at
the boundary is also directly ground. Gradual System~F
\cite{labrada2018gradualsystemf} and the systems studied by Igarashi et
al.~\cite{igarashi2017polymorphic} provide major polymorphic-gradual baselines,
including precision and graduality criteria that our explicit target does not
claim. Abstracting Gradual Typing \cite{garcia2016abstracting} and gradual type theory
\cite{new2019gradual,new2018gradualdt} give systematic accounts of gradual
languages and their metatheory.

\paragraph{Dependent interoperability.}
A related line studies safe interoperation between dependently typed and
simply typed components within one host language. Osera, Sj{\"o}berg, and
Zdancewic introduce dependent interoperability~\cite{osera2012dependent}, and
Dagand, Tabareau, and Tanter derive verified marshalling between indexed and
simple types from partial type equivalences, with dynamic checks at the
boundary~\cite{dagand2018interop}. Those systems connect two typing
disciplines inside a dependent host; we instead translate \emph{out of} the
dependent language altogether, into a standalone polymorphic blame calculus
with its own mechanized metatheory, and prove properties of the translation
rather than of a boundary combinator library.

\paragraph{Gradual dependent types.}
A line of work gradualizes dependent types directly: approximate normalization
\cite{eremondi2019approximate} and Gradual CIC \cite{lennon2022gradual} equip a
dependent theory with a dynamic type internally, and Tanter and Tabareau bring
gradual certified programming to Coq \cite{tanter2015gradualcertified}. These
make the \emph{source} gradual. We instead keep the source fully static (CoC) and
use an explicit dynamic/blame \emph{target}, with $\dyn$ absorbing dependency
that the polymorphic target cannot represent---a translation, not a gradual
surface-language design.
Cayenne \cite{augustsson1998cayenne} compiled a dependently typed language by
erasing to an untyped core; intrinsically typed interpreter interfaces (as
axiomatized in our \texttt{stlc.v}) are folklore in Agda
\cite{norell2008agda} and Idris. Our contribution is the extraction metatheory,
not an implementation of those examples.

%% ==========================================================================
\section{Conclusion}\label{sec:conclusion}
%% ==========================================================================

We have shown that extraction from a dependently typed language need not throw
away all types: with an explicit dynamic/blame target, a single verified translation preserves
the parametric-polymorphic skeleton of CoC programs, removes many dependent
indices from target types, and represents the translation's fallback cases---most visibly values used as
types---with $\dyn$ and, when forced operationally, reserved internal blame.  In this sense the work returns to Paulin-Mohring's
CoC-to-F$\omega$ vision~\cite{paulinmohring1989fomega}, but with a dynamic/blame
target and mechanized external-label safety.  The whole pipeline, including a
machine-checked F$\omega$-style polymorphic blame calculus, is formalized in
Rocq, with target progress/preservation and the extraction metatheorems---well-typed
extraction, syntactic simulation, and derivation independence---fully admit-free
(\S\ref{sec:mech}); a REPL with verified name
recovery makes its behavior
inspectable on signature-level examples. The current extractor does not implement Coq-style proof
erasure; proof terms remain ordinary residual terms. The results chart a design
point between fully typed and fully untyped extraction. Natural next steps are a
new typed logical relation and contextual-observation theorem, a transparent
algorithmic executable path, a proof-erasing optimization pass, and an
extension to inductive families and their large eliminations.

\section*{Data-Availability Statement}
The complete Rocq development, the extracted REPL, and the eighteen example
batches (with their checked-in golden output files) accompany this paper as an
artifact. The development builds with Rocq~9.0 and Dune; \texttt{dune test}
replays every golden output, and the assumption audit of \S\ref{sec:mech}
re-checks that the headline theorems depend only on \texttt{eq\_rect\_eq}.
The stock-extraction baseline of \S\ref{sec:examples} (the idiomatic-CIC
ports, the \texttt{Obj.magic} counts, and the segfaulting client) is
reproduced by \texttt{paper/stock-extraction/run.sh}.

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
